# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
The data has been loaded directly from provided file:
```{r}
dt <- read.csv("activity.csv", header = TRUE)
```

## What is mean total number of steps taken per day?
On the histogram below, there is total

```{r diagram1, fig.height=6, fig.width=6}
stps <- tapply(dt$steps,dt$date,sum)
hist(stps,col="blue",xlab="Steps per day", main="Total number of steps taken each day")
abline(v=median(stps,na.rm=T),col="red", lwd=2)
legend("topright", col="red", lwd=2, legend="mediane",bty="n")
rug(stps)
```
Most of the daily totals are concentrated under the bar in between 1000 and 1500 steps/day.

The histogram is slightly right skewed, but still resembling rather normal-distribution shape than a power-law one.

This bulk of the data can be easier observed on the boxplot diagram below:

```{r diagram2, fig.height=3, fig.width=6}
par(mar=c(2, 2, 2, 2))
boxplot(stps,horizontal=T,col="blue")
```

The same, more precisely can be seen in the 5-number summary:
```{r}
summary(stps)
```
The **median**, is practicaly equal to the **mean** which is = `r format(mean(stps, na.rm=T),digits=7)`

There are very few (but maybe interesting to look closer at) outliers and there are 8 days rejected out of `r length(unique(dt$date))` because of missing values for these days.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
