# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
This analysis makes use of data from a personal activity monitoring device. 
The device collects data at 5 minute intervals through out the day. 

The data consists of two months of data from an anonymous individual collected between 1-st of October and 30-th of November 2012, the number of steps taken in 5 minute intervals each day.

**Source of the data**: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

**Dataset**: contains a total of 17,568 observations.

The variables included in the dataset are:

1. *steps*: Number of steps taking in a 5-minute interval (there are some missing values)
2. *date*: The date on which the measurement was taken in YYYY-MM-DD format
3. *interval*: Identifier for the 5-minute interval in which measurement was taken

The data has been loaded in R using following code:
```{r}
dt <- read.csv("activity.csv", header = TRUE)
stps <- tapply(dt$steps,dt$date,sum)
```

## What is mean total number of steps taken per day?
The **mean total number of steps taken per day** is = `r format(mean(stps, na.rm=T),digits=7)`,
to answer this question a histogram was created:

```{r diagram1, fig.height=6, fig.width=6}
stps <- tapply(dt$steps,dt$date,sum)
hist(stps,col="blue",xlab="Steps per day", main="Total number of steps taken each day")
abline(v=median(stps,na.rm=T),col="red", lwd=2)
legend("topright", col="red", lwd=2, legend="mediane=mean",bty="n")
rug(stps)
```

The histogram is normal-distribution shaped and slightly right skewed.
Bulk of the daily totals is under the bar in between 1000 and 1500 steps/day.

The concentration of the data can be easier observed on the boxplot diagram below:

```{r diagram2, fig.height=3, fig.width=6}
par(mar=c(2, 2, 2, 2))
boxplot(stps,horizontal=T,col="blue")
```

..or, more precisely in the 5-number summary:
```{r}
summary(stps)
```
The **median**, is practically equal to the **mean** which is = `r format(mean(stps, na.rm=T),digits=7)`

There are very few (but maybe interesting to look closer at) outliers and there were 8 days excluded from observations out of `r length(unique(dt$date))` because of missing values for these days.

## What is the average daily activity pattern?
By calculating totals of of steps taken, averaged across all days the daily activity pattern of the person can be observed:

```{r}
par(mfrow = c(1, 1), mar = c(4, 4, 2, 1), oma = c(2, 0, 2, 0))
pattrns <- tapply(dt$steps,dt$interval,mean, na.rm=T)
plot(as.numeric(names(pattrns)),pattrns,col="red",type="l",ylab="avg. steps/5min",xlab="time (hhmm)",lab=c(5,10,7))
title("Average daily activity pattern")

maxActivityTime <- as.integer(names(which.max(pattrns)))
sprintf("%d:%d",maxActivityTime%/%100,maxActivityTime%%100)
abline(v=maxActivityTime,col="green",lwd=2)
legend("topright", col="green", lwd=2, legend="max activity",bty="n")
```
On average, the person's activity starts about 5:30 and ceased at almost midnight, with the maximum at 8:35.

## Imputing missing values

As mentioned earlier the data does not cover all of time-intervals between 12/10/01 and 12/11/30. 
Actually there are `r format(sum(is.na(dt$steps)),digits=7)` of rows missing in the dataset:
```{r}
nulls <- is.na(dt$steps); length(nulls)
```
To deal with missing values a simple strategy for filling in all of the missing values has been implemented to create the new dataset without missing values:

```{r}
dtNew <- dt
dtNew$steps <- ifelse(nulls, as.integer(pattrns[as.character(dt$interval)]),dt$steps )
```
It replaces missing values in *steps* field at particular time-interval with  averages for this very time-interval across all days.

To see how it changed the distribution of values let's see the histogram for total number of steps taken each day:

```{r}
stpsNew <- tapply(dtNew$steps,dtNew$date,sum)
hist(stpsNew,col="red",xlab="Steps per day", main="Total number of steps taken each day")
rug(stps)

```

and comparisson of 5-number summaries for both datasets:

```{r}
summary(stps)
summary(stpsNew)
```

There is not much change in between both histograms, except that the new one is *narrower* or *steeper*, i.e. there are more measurements in the middle of the range then it was, and variance seems to have decreased. But the range and the mean/median remained almost unchanged.


## Are there differences in activity patterns between weekdays and weekends?
There are interesting differences in the daily patterns of activity to be observed when looking at  different days of a week. The difference is particulary visible in between weekday days and Sundays:

```{r}

dtNew$wday <- ifelse(weekdays(as.Date(dtNew$date))!="Sunday", "weekday","weekend")
tail(dtNew$wday)
dtNew$wday <-as.factor(dtNew$wday)

pattrnsW <- tapply(dtNew[dtNew$wday=="weekday",]$steps,dtNew[dtNew$wday=="weekday",]$interval,mean)
pattrnsS <- tapply(dtNew[dtNew$wday=="weekend",]$steps,dtNew[dtNew$wday=="weekend",]$interval,mean)

par(mfrow = c(2, 1), mar = c(0, 4, 0, 1), oma = c(2, 0, 2, 0))
plot((names(pattrnsW)),pattrnsW,col="red",type="l",lab=c(1,5,7),ylab="avg. steps/5min",xlab="")
legend("bottomleft", legend="weekday",bty="n")
plot(as.numeric(names(pattrnsS)),pattrnsS,col="red",type="l",lab=c(5,5,7),ylab="avg. steps/5min",xlab="time (5 min intervals)")
legend("bottomleft", legend="Sunday",bty="n")
mtext("  Comparison of day activity patterns", outer = TRUE)
```

The differences in activity patterns are more prominent than in between weekend (Sat-Sun) and weekday (Mon-Fri)
```{r}

dtNew$wday[as.Date(dtNew$date)=="Saturday"] <- "weekend"

pattrnsW <- tapply(dtNew[dtNew$wday=="weekday",]$steps,dtNew[dtNew$wday=="weekday",]$interval,mean)
pattrnsS <- tapply(dtNew[dtNew$wday=="weekend",]$steps,dtNew[dtNew$wday=="weekend",]$interval,mean)

par(mfrow = c(2, 1), mar = c(0, 4, 0, 1), oma = c(2, 0, 2, 0))
plot((names(pattrnsW)),pattrnsW,col="red",type="l",lab=c(1,5,7),ylab="avg. steps/5min",xlab="")
legend("bottomleft", legend="weekday",bty="n")
plot(as.numeric(names(pattrnsS)),pattrnsS,col="red",type="l",lab=c(5,5,7),ylab="avg. steps/5min",xlab="time (5 min intervals)")
legend("bottomleft", legend="Sunday",bty="n")
mtext("  Comparison of day activity patterns", outer = TRUE)
```



